---
title: "Vignette for mle package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette for mle package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## Sections

1. Introduction
2. Maximum likelihood estimation (MLE)
3. Using R's optimization routines
4. Installation

## Introduction

### Maximum likelihood estimation (MLE)

Forthcoming

### Using R's optimization routines

R has several options for general purpose optimization. Furthermore, there are a wide variety of R packages that have been created that provide very  powerful optimization tools for a variety of problems; the package here uses R's base functions primarily for ease of application. This package focuses on two functions, `optim` and `nlm`.

The `optim` function is based on code originally from Nash (1990), and provides a range of optimization algorithms for attempting to find either global maxima or minima. As an example, suppose we want to use `optim` to obtain maximum likelihood estimates when fitting the normal distribution to a set of data. First, we will simulate some data to fit:

```{r}
# Define a set of generating parameters
gp = c( mu = 100, sigma = 15 )

# Generate a set of data
set.seed = 1
x = rnorm( 100, gp[1], gp[2] )
```

To use `optim` to obtain the maximum likelihood estimates, we must provide a set of starting values to initialize the parameter search, and then we must provide a function that takes the data and computes the sum of the log-likelihoods:

```{r}
# Starting values
start_val = c( 80, 30 )

# Compute the sum of the log-likelihoods for the normal distribution
fn = function( prm, x ) {
  # prm = A vector of parameters; the mean and standard deviation
  # x = A vector of data
  
  sll = sum( dnorm( x, prm[1], prm[2], log = T ) )
  
  return( sll )
}
```

Finally, we must set `optim` to attempt to maximize the function instead of minimize:

```{r}
control = list( fnscale = -1 )
```

We can now make the call to `optim`:

```{r}
estimates = optim( start_val, fn, x = x, control = control )
```

Note that a warning is generated - this is because the parameter $\sigma$ for the normal distribution must be positive, but optimization routine can propose negative values for this parameter.

Due to the simplicity of the model, we can compare the estimates returned by `optim` to the maximum likelhood estimates based on analytic method:
```{r}
print( "Estimates (optim)" )
print( round( estimates$par, 2 ) )

print( "True maximum likelihood estimates" )
print( round( c( mean(x), sd(x) ), 2 ) )
```

As can be seen, the estimate for the mean from `optim` is equivalent to the true estimate, but the estimate for the standard deviation is slightly off.

## Installation

The package can be installed via "devtools":

```{r,eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("rettopnivek/mle")
```

Once the package has been installed, we can load it for easy use:

```{r}
library(mle)
```

## The likelihood function

The 'mle' package provides a template for creating functions to compute the sum of the log-likelihoods of a given model that are compatible with the package's convenience functions. This template can be used by calling `mle_fn_template()` and copying and modifying the resulting output:

```{r}
mle_fn_template()
```

While the function does not have to be named `mle_fn`, it must take four parameters: `prm`, `dat`, `sum`, and `priors`. Furthermore, for compatibility with the S3 class ??? options, the function must return the sum of the log-likelihoods if `sum = T` or a vector of the log-likelihoods for every observation if `sum = F`.

## ???

```{r}
mle_fn = function( prm, dat, 
                   sum = T, priors = NULL ) { 
  # Initial setup 
  y = dat; n = 1 # Extract data
  theta = 1/(1 + exp( -prm ) ) # Bound between 0 and 1

  # Calculate the log-likelihoods 
  ll = dbinom( y, n, theta, log = T )
  if ( !sum ) return( ll )

  # Sum the log-likelihoods 
  sll = sum( ll ) 

  # Incorporate priors (Optional) 
  if ( !is.null(priors) ) { 
    sll = sll +  
    dbeta( theta, priors[1], priors[2], log = T )
  } 

  # Check for NA values 
  if ( is.na( sll ) ) sll = -Inf 

  return( sll ) 
}
```
